From correlation to causality: statistical approaches to learning regulatory relationships in large-scale biomolecular investigations

### 2015-12-18 manuscript/write_up.tex
tweaked responses

We illustrate this problem with a computer simulation, inspired by Fan et. al\cite{fan2014challenges} but translated to our context. First, we simulated an experiment that quantifies the abundances of 60 proteins in 60 biological samples. Second, we simulated another experiment where the number of proteins was increased from 60 to 6000. In both experiments the proteins are completely independent from each other, and each protein in each replicate is assigned a value randomly drawn from a Gaussian distribution. In other words, we do not expect any biologically meaningful associations in these data. We repeated each of these simulations 500 times. Figure \ref{spurious_corr.png} shows for each experiment the histogram of the highest Pearson correlation across any pair of proteins in the 500 instances of the simulation. As can be seen, the increase to 6000 proteins produces relatively large maximum pairwise correlations, demonstrating that an increase in the number of proteins leads to an increase in spurious correlation.  This is clearly a problem when high Pearson correlation is used as an initial evidence of biological function.

@@

\added{To build dependence into the simulated data, we created multiple sets of 3-proteins "triplets" A, B, and C, where all three are dependent but A and B are conditionally independent given C.  Thus perform perfectly on the $i^{th}$ triplet, the algorithm would need two true positives (rejecting conditional independence between $A_i$ -- $C_i$ and $B_i$ -- $C_i$), find the $A_i$ and $B_i$ pair conditionally independent, and find any pair containing a node in the $i^th$ triplet and any node not in the $i^th$ triplet to be independent.  We made it such that the proteins in each triplet generate measurements with Gaussian values with mean 0 and variance 1, and that the covariance between A--C and B--C was .3.  We simulated data from a multivariate Gaussian. }

\added{Again we run the simulation 500 times, in each iteration we performed step 1 by calculating pairwise Pearson correlations, and then Step 2 by applying the Grow-Shrink algorithm \cite{margaritis2003learning} for causal inference, an instance from a class of algorithms called constraint-based network inference algorithms\cite{spirtes2000causation} that proceed with a series of conditional independence tests that control the false positive rate.  To assure we weren't observing algorithm specific effects, we also repeated the simulation other algorithms from this class\cite{tsamardinos2003algorithms}\cite{yaramakala2005speculative}.  The results did not differ (results not shown).}

>>>>

We illustrate this problem with a computer simulation, inspired by Fan et. al\cite{fan2014challenges} but translated to our context. First, we simulated an experiment that quantifies the abundances of 60 proteins in 60 biological samples. Second, we simulated another experiment where the number of proteins was increased from 60 to 6000. In both experiments the proteins are completely independent from each other, and each protein in each replicate is assigned a value randomly drawn from a Gaussian distribution. In other words, we do not expect any biologically meaningful associations in these data. We repeated each of these simulations 500 times. Figure \ref{spur_corr} shows for each experiment the histogram of the highest Pearson correlation across any pair of proteins in the 500 instances of the simulation. As can be seen, the increase to 6000 proteins produces relatively large maximum pairwise correlations, demonstrating that an increase in the number of proteins leads to an increase in spurious correlation.  This is clearly a problem when high Pearson correlation is used as an initial evidence of biological function.

@@

\added{To build dependence into the simulated data, we created multiple sets of 3-proteins "triplets" A, B, and C, where all three are dependent but A and B are conditionally independent given C.  Thus perform perfectly on the $i^{th}$ triplet, the algorithm would need two true positives (rejecting conditional independence between $A_i$ -- $C_i$ and $B_i$ -- $C_i$), find the $A_i$ and $B_i$ pair conditionally independent, and find any pair containing a node in the $i^th$ triplet and any node not in the $i^{th}$ triplet to be independent.  We made it such that the proteins in each triplet generate measurements with Gaussian values with mean 0 and variance 1, and that the covariance between A--C and B--C was .3.  We simulated data from a multivariate Gaussian. }

\added{Again we run the simulation 500 times, in each iteration we performed step 1 by calculating pairwise Pearson correlations, and then Step 2 by applying the Grow-Shrink algorithm \cite{margaritis2003learning} for causal inference, an instance from a class of algorithms called constraint-based network inference algorithms\cite{spirtes2000causation} that proceed with a series of conditional independence tests that control the false positive rate.  To assure we weren't observing algorithm specific effects, we also repeated the simulation other algorithms from this class\cite{tsamardinos2003algorithms}\cite{yaramakala2005speculative} .  The results did not differ (results not shown).}
